How did local health infrastructure and socio-political factors within different states and counties in the United States affect the disparities in COVID-19 outcomes, and what lessons can be learned for more targeted public health preparedness and response strategies in future pandemics?

### 2024-02-13 outputs/paper/paper.tex
new table, new references, lots of other changes

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\KOMAoption{captions}{tableheading}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={How did local health infrastructure and socio-political factors within different states and counties in the United States affect the disparities in COVID-19 outcomes, and what lessons can be learned for more targeted public health preparedness and response strategies in future pandemics?},
  pdfauthor={Adrian Ly; Sakhil Goel; Hannah Yu},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{How did local health infrastructure and socio-political factors
within different states and counties in the United States affect the
disparities in COVID-19 outcomes, and what lessons can be learned for
more targeted public health preparedness and response strategies in
future pandemics?\thanks{Code and data are available at:
https://github.com/hannahyu07/US-Covid-Analysis.git}}
\author{Adrian Ly \and Sakhil Goel \and Hannah Yu}
\date{February 12, 2024}

\begin{document}
\maketitle
\begin{abstract}
First sentence. Second sentence. Third sentence. Fourth sentence.
\end{abstract}

\renewcommand*\contentsname{Table of contents}
{
\hypersetup{linkcolor=}
\setcounter{tocdepth}{3}
\tableofcontents
}
\section{Introduction}\label{introduction}

@@

\emph{The introduction is self-contained and tells a reader everything
they need to know including: 1) broader context to motivate; 2) some
detail about what the paper is about; 3) a clear gap that needs to be
filled; 4) what was done; 5) what was found; 6) why it is important; 7)
the structure of the paper. A reader should be able to read only the
introduction and know what was done, why, and what was found. Likely 3
or 4 paragraphs, or 10 per cent of total. The estimand is clearly stated
in the introduction.}

\section{Data}\label{sec-data}

Our results are summarized in (\textbf{fig:fig-GHS?}).

\subsection{Source}\label{source}

The datasets utilized in this paper were mainly obtained from the
\texttt{original\ paper} (Nuzzo and Ledesma 2023). Additionally, to
address the original paper's lack of US Covid statistics and political
party support data, we incorporated information from Jack and Oster
(2023) and Elflein (2023).

Jack and Oster (2023) discusses the long-term impacts of COVID-related
school closures. From this source, we utilized the dataset on voting
shares during the 2020 election by county. Elflein (2023) provides
summaries of COVID-19 death rates in the United States as of March 2023,
organized by state. Analyzing results from both datasets allows us to
explore the relationship between political affiliation and COVID-19
outcomes.Our reproduction aims to fill these gaps and also includes
tables and graphs that were not presented in the original paper to
support our findings.

\subsection{Methodology}\label{methodology}

\texttt{R} (R Core Team 2022) was the language and environment used for
the bulk of this analysis, alongside \texttt{tidyverse} (Wickham et al.
2019), \texttt{sf} (Pebesma 2018), \texttt{readxl} (Wickham and Bryan
2023), \texttt{knitr} (Xie 2014), \texttt{janitor} (Firke 2023),
\texttt{lubridate} (Grolemund and Wickham 2011), \texttt{dplyr} (Wickham
et al. 2023), \texttt{data.table} (Barrett et al. 2024),
\texttt{RColorBrewer} (Neuwirth 2022), \texttt{ggpubr} (Kassambara
2023), \texttt{ggplot2} (Wickham 2016), \texttt{here} (Müller 2020),
\texttt{kableExtra} (Zhu 2024), \texttt{webshot} (Chang 2023a),
\texttt{webshot2} (Chang 2023b), \texttt{gt} (Iannone et al. 2024), and
\texttt{scales} (Wickham, Pedersen, and Seidel 2023).

\subsection{Data Measurement}\label{data-measurement}

\emph{A thorough discussion of measurement, relating to the dataset, is
provided in the data section.}

\subsection{Data cleaning}\label{data-cleaning}

@@

\centering{

\includegraphics{paper_files/figure-pdf/fig-GHS-1.pdf}

@@

\end{figure}%

\section{Results}\label{results}

Our results are summarized in the following figures. \textbf{?@fig-ELE}
illustrates the trend of life expectancy at birth across different
racial groups over time. Table~\ref{tbl-life_exp} provides a more
detailed breakdown of the life expectancy before and during COVID
for\ldots{} Unfortunately, due to the unavailability of data over time
for Asians and American Indian and Alaska Native communities, they have
been excluded from the time series graph.

An intriguing observation is the consistently higher life expectancy
among Hispanic individuals compared to other groups, even amidst the
challenges posed by COVID-19. On the other hand, Black individuals l
have consistently exhibited lower life expectancy, which further
declined notably in 2021, reaching just over 70.8 years old. The life
expectancy trends of white people and all other races and origins remain
close together throughout the 14 year time period, with minimum
variance. Specifically, the life expectancy for White individuals
decreased from 78.8 to 76.4 years old from pre-pandemic levels in 2019
to 2021, while for Hispanic individuals, it dropped by 4.2 years, and
for Black individuals, it declined by 4 years during the same period.

\begin{table}

\caption{\label{tbl-life\_exp}Life Expectancy by Race (2019-2021)}

\centering{

@@

\end{table}%

@@

\centering{

\includegraphics{paper_files/figure-pdf/fig-CLE-1.pdf}

@@

\end{figure}%

Figure~\ref{fig-CLE} serves as a valuable addition to our previous
analysis with the inclusion of Asians and American Indians and Alaska
Natives during the critical period from 2019 to 2021. With these
additional ethnic categories, we can discern that American Indians and
Alaska Natives experienced significant impacts from the pandemic, with a
decrease of 4.7 years in the first year and 1.9 years in the second
year, totaling 6.6 years.

With a chart that depicts the change of life expectancy each year by
ethnicity, we are able to better gain a more comprehensive understanding
of the impact of COVID by ethnicity. While the declines in life
expectancy are smaller in magnitude from 2020 to 2021, they are notably
minimal for Asians and Hispanics, with reductions of -0.1 and -0.2 years
respectively. (Insert some sources tmrw)

Numerous sources have indicated a correlation between a state's
political affiliation and its handling of COVID issues (sources to be
inserted tomorrow). To testify the claim, we collected the voting data
for each of the 50 states for the 2020 presidential election. We then
selected the ten states with the highest proportions of Republican
votes.These voting patterns are visualized in Figure
Figure~\ref{fig-Vote}, where red symbols represent Republican votes and
blue symbols represent Democratic votes. Among the top ten states,
Tennessee recorded the highest total number of votes, while Wyoming
boasted the highest proportion of Republican votes.

@@

\centering{

\includegraphics{paper_files/figure-pdf/fig-Vote-1.pdf}

@@

\end{figure}%

Given the variation in population sizes among states, direct comparisons
of COVID case numbers are inherently flawed. As an alternative, we
employed death rates per 100,000 people as a metric for evaluating each
state's COVID preparedness and situation. We then produced
Table~\ref{tbl-deaths} that ranks the top ten states with the highest
death rates from COVID per 100,000 people to examine the potential
correlation between party preferences and COVID related deaths. Notably,
We found that six of the ten top Republican states made a reappearance
in the top death rates table; these states are Oklahoma, West Virginia,
Arkansas, Alabama, Tennessee, and Kentucky. (Insert quotes about
republican party and covid tmrw)

\begin{table}

\caption{\label{tbl-deaths}Top 10 States with Highest Death Rates from
COVID-19 (per 100,000 people)}

\centering{

>>>>

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{caption}
\KOMAoption{captions}{tableheading}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={How did local health infrastructure and socio-political factors within different states and counties in the United States affect the disparities in COVID-19 outcomes, and what lessons can be learned for more targeted public health preparedness and response strategies in future pandemics?},
  pdfauthor={Adrian Ly; Sakhil Goel; Hannah Yu},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{How did local health infrastructure and socio-political factors
within different states and counties in the United States affect the
disparities in COVID-19 outcomes, and what lessons can be learned for
more targeted public health preparedness and response strategies in
future pandemics?\thanks{Code and data are available at:
https://github.com/hannahyu07/US-Covid-Analysis.git}}
\author{Adrian Ly \and Sakhil Goel \and Hannah Yu}
\date{February 13, 2024}

\begin{document}
\maketitle
\begin{abstract}
First sentence. Second sentence. Third sentence. Fourth sentence.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, frame hidden, enhanced, interior hidden, sharp corners]}{\end{tcolorbox}}\fi

\renewcommand*\contentsname{Table of contents}
{
\hypersetup{linkcolor=}
\setcounter{tocdepth}{3}
\tableofcontents
}
\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

@@

Figure~\ref{fig-GHS} allows us to visualize the Global Health Security
(GHS) Index which is designed to assess a country's capabilities to
prevent, detect, and respond to significant infectious disease
outbreaks. A higher Global Health Security (GHS) Index score indicates a
stronger health system capacity for pandemic readiness. The GHS Index
assesses the readiness of countries for pandemics and other significant
biological threat emergencies, using a comprehensive framework that
includes measurements grouped into six broad categories: prevention,
detection, response, health system capacity, compliance with
international norms, and risk environment. Higher scores suggest that a
country has a more robust health system capacity, with better
capabilities to prevent, detect, respond to, and mitigate the spread of
an epidemic.

Conversely, a lower GHS Index score implies weaker health system
capacities, suggesting that the country may be less prepared to handle
pandemics and other significant biological threats. Lower scores
indicate deficiencies in one or more of the critical areas assessed by
the Index, which could potentially hinder the country's ability to
effectively manage and control the spread of infectious diseases.

@@

{\centering \includegraphics{paper_files/figure-pdf/fig-GHS-1.pdf}

@@

\end{figure}

\hypertarget{sec-data}{%
\section{Data}\label{sec-data}}

\hypertarget{source}{%
\subsection{Source}\label{source}}

The datasets utilized in this paper were mainly obtained from the
\texttt{original\ paper} (Nuzzo and Ledesma 2023). The data used by the
paper was compiled from various sources such as the Arias et al. (2022),
Arias and Xu (2022), Bell and Nuzzo (2021), Systems Science and
Engineering (2023), Dong, Du, and Gardner (2020), \emph{COVID-19 Excess
Mortality Estimates 2020-2021} (2022), Ledesma et al. (2023), Health
Statistics (2022), {``Death Rate, Crude (Per 1,000 People)
(SP.DYN.CDRT.IN)''} (2022), Economic and Division (2022), and Wang et
al. (2022). Additionally, to address the original paper's lack of US
COVID statistics and political party support data, we incorporated
information from Jack and Oster (2023) and (Elflein 2023). The political
party support data we utilized from Jack and Oster (2023) was compiled
from McGovern (2009-\/-2020).

Jack and Oster (2023) discusses the long-term impacts of COVID-related
school closures. From this source, we utilized the dataset on voting
shares during the 2020 election by county. Elflein (2023) summarizes
COVID-19 death rates in the United States as of March 2023, organized by
state. Analyzing results from both datasets allows us to explore the
relationship between political affiliation and COVID-19 outcomes. Our
reproduction aims to fill these gaps and includes tables and graphs not
presented in the original paper to support our findings.

\hypertarget{methodology}{%
\subsection{Methodology}\label{methodology}}

This paper will replicate the data that was originally presented by
Nuzzo and Ledesma (2023), as previously mentioned. We then transform the
replication to our data analysis incorporating the COVID statistics and
political party support data. We seek to use our initial replication and
following own data analysis to provide a fresh part of the original
paper, enhancing the with more data. Different from the original paper
which spent a lot of time talking about how badly prepared the US was
against other countries, We also specifically diverge our focus to the
political and socioeconomic factors that impacted the US's COVID
outcome.

\texttt{R} (R Core Team 2022) was the language and environment used for
the bulk of this analysis, alongside \texttt{tidyverse} (Wickham et al.
2019), \texttt{sf} (Pebesma 2018), \texttt{readxl} (Wickham and Bryan
2023), \texttt{knitr} (Xie 2014), \texttt{janitor} (Firke 2023),
\texttt{lubridate} (Grolemund and Wickham 2011), \texttt{dplyr} (Wickham
et al. 2023), \texttt{data.table} (Barrett et al. 2024),
\texttt{RColorBrewer} (Neuwirth 2022), \texttt{ggpubr} (Kassambara
2023), \texttt{ggplot2} (Wickham 2016), \texttt{here} (Müller 2020),
\texttt{kableExtra} (Zhu 2024), \texttt{webshot} (Chang 2023a),
\texttt{webshot2} (Chang 2023b), \texttt{gt} (Iannone et al. 2024) and
\texttt{scales} (Wickham, Pedersen, and Seidel 2023).

\hypertarget{data-cleaning}{%
\subsection{Data cleaning}\label{data-cleaning}}

From all the data provided from Nuzzo and Ledesma (2023), Jack and Oster
(2023), and Elflein (2023), we mainly utilize Bell and Nuzzo (2021),
Health Statistics (2022), McGovern (2009-\/-2020), and Elflein (2023).
The data from Nuzzo and Ledesma (2023) is well-organized and require
minimal cleaning for analysis.

The data collected from McGovern (2009-\/-2020) is at the county level
as depicted on Table~\ref{tbl-sample}. However, since our analysis
focuses on people's political preferences at the state level, we create
a new data frame called ``vote\_data\_by\_state'' to aggregate voting
data by state. We utilize the ``group\_by'' function to generate new
variables at the state level, calculating the total votes, total
Republican party votes, total Democratic party votes, and the total vote
difference between the two parties by county. Additionally, We also
create variables for the average percentage of votes each party receives
and the difference by each state.

To determine the total number of votes in the Top 10 states with the
highest proportion of Republican votes, we create another new data frame
called ``sorted\_states'' from the ``vote\_data\_by\_state'' data frame.
We include a new variable ``proportion\_republican'' that calculates the
proportion of votes the Republican party receives from each state and
orders the states by the highest proportion of Republican votes.

\hypertarget{tbl-sample}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 18\tabcolsep) * \real{0.0791}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 18\tabcolsep) * \real{0.0863}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 18\tabcolsep) * \real{0.1079}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 18\tabcolsep) * \real{0.0719}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 18\tabcolsep) * \real{0.0719}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 18\tabcolsep) * \real{0.0863}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 18\tabcolsep) * \real{0.0791}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 18\tabcolsep) * \real{0.1079}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 18\tabcolsep) * \real{0.1079}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 18\tabcolsep) * \real{0.2014}}@{}}
\caption{\label{tbl-sample}Sample of Vote Share Data}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
State Name
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
County FIPS
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
County Name
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Votes GOP
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Votes DEM
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Total Votes
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Difference
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Percentage GOP
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Percentage DEM
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Percentage Point Difference
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
State Name
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
County FIPS
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
County Name
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Votes GOP
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Votes DEM
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Total Votes
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Difference
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Percentage GOP
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Percentage DEM
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Percentage Point Difference
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Alabama & 1001 & Autauga County & 19838 & 7503 & 27770 & 12335 &
0.7143680 & 0.2701837 & 0.4441844 \\
Alabama & 1003 & Baldwin County & 83544 & 24578 & 109679 & 58966 &
0.7617137 & 0.2240903 & 0.5376234 \\
Alabama & 1005 & Barbour County & 5622 & 4816 & 10518 & 806 & 0.5345123
& 0.4578817 & 0.0766305 \\
Alabama & 1007 & Bibb County & 7525 & 1986 & 9595 & 5539 & 0.7842626 &
0.2069828 & 0.5772798 \\
Alabama & 1009 & Blount County & 24711 & 2640 & 27588 & 22071 &
0.8957155 & 0.0956938 & 0.8000217 \\
\end{longtable}

The data from Arias and Xu (2022) lacks column names for each column. To
retrieve data by column, we assign column names ``State'' and
``Death\_Rate'' to the variables of interest. Subsequently, to find the
top 10 states with the highest death rates, we rank the states based on
death rate. Lastly, to create the final figure that incorporates the
COVID death rate and people's political preferences by state, we merge
the two datasets from McGovern (2009-\/-2020) and Elflein (2023) and
proceed with our analysis.

\hypertarget{data-measurement}{%
\subsection{Data Measurement}\label{data-measurement}}

\hypertarget{measurement-techniques}{%
\subsubsection{Measurement Techniques}\label{measurement-techniques}}

Manual Data Extraction involved careful review of reports and tables to
ensure accurate data capture, particularly for datasets like life
expectancy where this meticulous approach was paramount; this was
coupled with Standardization and Adjustment processes where excess death
data and age-standardized excess death rates underwent significant
processing, including the comparison of observed deaths during the
pandemic to expected deaths based on historical trends, with adjustments
made for the age distribution of the population; concurrently,
Geospatial Data Handling was employed for mapping the GHS Index scores,
utilizing Geographic Information Systems (GIS) to visually represent
pandemic preparedness across various countries, and this was
complemented by Real-Time Data Aggregation, which entailed aggregating
COVID-19 death data in real-time from authoritative sources such as the
Johns Hopkins University Dashboard, necessitating continuous data
monitoring and updates to maintain accuracy and relevance.

\hypertarget{adjustments-and-considerations}{%
\subsubsection{Adjustments and
Considerations}\label{adjustments-and-considerations}}

Accounting for underreporting through the measurement of excess deaths
provides a more comprehensive view of the pandemic's impact by capturing
deaths that may have been missed or misclassified as COVID-19, while the
process of age standardization adjusts these excess death rates to a
standard age distribution, enabling meaningful comparisons between
countries with varying demographic profiles. However, the analysis had
to navigate challenges associated with varying data quality and
availability across different countries, especially given the diverse
COVID-19 reporting standards and testing capacities. Moreover, the study
also acknowledges the significant influence of political and social
factors on the pandemic's outcomes, which, despite being more
challenging to quantify, are essential for a holistic understanding of
the pandemic's impact.

\hypertarget{results}{%
\section{Results}\label{results}}

Our results are summarized in the following figures.
Figure~\ref{fig-ELE} illustrates the trend of life expectancy at birth
across different racial groups over time. Unfortunately, due to the
unavailability of data over time for Asian, American Indian, and Alaska
Native communities, they have been excluded from the time series graph.
Table~\ref{tbl-life-exp} provides a more detailed breakdown of the life
expectancy before and during COVID for Asian, American Indian, Alaska
Native and other racial communities.

An intriguing observation from Figure~\ref{fig-ELE} is the consistently
higher life expectancy among Hispanic individuals compared to other
groups, even amidst the challenges posed by COVID-19. On the other hand,
Black individuals have consistently exhibited lower life expectancy,
which further declined notably in 2021, reaching just over 70.8 years
old. The life expectancy trends of white people and all other races and
origins remain close together throughout the 14 years, with minimum
variance.

While all racial groups experienced declines in their average life
expectancy, the declines vary greatly. The life expectancy for White
individuals decreased from 78.8 to 76.4 years old from pre-pandemic
levels in 2019 to 2021, while for Hispanic individuals, it dropped by
4.2 years, and for Black individuals, it declined by 4 years during the
same period. Our data aligns with the findings of Andrasfay and Goldman
(2022), which states that the Black and Hispanic populations experienced
twice as much life expectancy decrease as the white population.''

{\centering \includegraphics{paper_files/figure-pdf/fig-ELE-1.pdf}

@@

\caption{\label{fig-ELE}Estimates of Life Expectancy at Birth, by Race
2006-2021}

\end{figure}

@@

{\centering \includegraphics{paper_files/figure-pdf/fig-CLE-1.pdf}

@@

\end{figure}

@@

{\centering \includegraphics{paper_files/figure-pdf/fig-Vote-1.pdf}

@@

\end{figure}

\hypertarget{tbl-deaths}{}
\begin{longtable}{rlr}
\caption{\label{tbl-deaths}Top 10 States with Highest Death Rates from COVID-19 (per 100,000
people) }\tabularnewline

\toprule
Rank & State & Deaths \\ 
\midrule\addlinespace[2.5pt]
1 & Arizona & 455 \\ 
2 & Oklahoma & 454 \\ 
3 & Mississippi & 449 \\ 
4 & West Virginia & 444 \\ 
5 & New Mexico & 432 \\ 
6 & Arkansas & 431 \\ 
7 & Alabama & 429 \\ 
8 & Tennessee & 428 \\ 
9 & Michigan & 423 \\ 
10 & Kentucky & 406 \\ 
\bottomrule
\end{longtable}

\hypertarget{tbl-life-exp}{}
\begin{longtable}{rrrrrrr}
\caption{\label{tbl-life-exp}Life Expectancy by Race (2019-2021) }\tabularnewline

\toprule
Year & All races and origins & Hispanic & AIAN & Asian & Black & White \\ 
\midrule\addlinespace[2.5pt]
2019 & 78.8 & 81.9 & 71.8 & 85.6 & 74.8 & 78.8 \\ 
2020 & 77.0 & 77.9 & 67.1 & 83.6 & 71.5 & 77.4 \\ 
2021 & 76.1 & 77.7 & 65.2 & 83.5 & 70.8 & 76.4 \\ 
\bottomrule
\end{longtable}

Figure~\ref{fig-CLE} serves as a valuable addition to our previous
analysis with the inclusion of Asians, American Indians, and Alaska
Natives during the critical period from 2019 to 2021. With these
additional ethnic categories, we can discern that American Indians and
Alaska Natives experienced significant impacts from the pandemic, with a
decrease of 4.7 years in the first year and 1.9 years in the second
year, totaling 6.6 years.

With a chart that depicts the change in life expectancy each year by
ethnicity, we can better gain a more comprehensive understanding of the
impact of COVID by ethnicity. While the declines in life expectancy are
smaller in magnitude from 2020 to 2021, they are notably minimal for
Asians and Hispanics, with reductions of -0.1 and -0.2 years
respectively.

Various studies have confirmed the existence of correlation between a
state's political affiliation and its handling of COVID issues (Datz
(2022)). To testify to the claim, we collected the voting data for each
of the 50 states for the 2020 presidential election. We then selected
the ten states with the highest proportions of Republican votes. These
voting patterns are visualized in Figure Figure~\ref{fig-Vote}, where
red symbols represent Republican votes and blue symbols represent
Democratic votes. Among the top ten states, Tennessee recorded the
highest total number of votes, while Wyoming boasted the highest
proportion of Republican votes.

Given the variation in population sizes among states, direct comparisons
of COVID-19 case numbers are inherently flawed. As an alternative, we
employed death rates per 100,000 people as a metric for evaluating each
state's COVID preparedness and situation. We then produced
Table~\ref{tbl-deaths} that rank the top ten states with the highest
death rates from COVID-19 per 100,000 people to examine the potential
correlation between party preferences and COVID-related deaths. Notably,
We found that six of the ten top Republican states made a reappearance
in the top death rates table; these states are Oklahoma, West Virginia,
Arkansas, Alabama, Tennessee, and Kentucky. The higher mortality rates
observed in Republican-leaning states may be attributed to their
preference for less stringent measures compared to Democratic-leaning
states, which tend to favor stricter measures (VanDusky-Allen and
Shvetsova (2021)).

Following our previous analysis regarding individuals' political
affiliations, we have developed Figure
Figure~\ref{fig-deaths-rates-votes}, which encompasses all 50 states of
the US along with their political leanings based on which party garnered
the majority votes. This information is juxtaposed against their
respective COVID-19 death rates.

While we cannot make any definitive assertions about stark differences,
we do observe that the Republican-leaning states are slightly more
clustered around higher death rates ranging from 350 to 450 deaths,
while the Democratic-leaning states appear to be more evenly
distributed, and notably one Democratic-leaning state has the lowest
death rate. An intriguing observation is that although many
Republican-leaning states demonstrate higher COVID death rates, it is
noteworthy that Arizona, typically considered a Democratic-leaning
state, records the highest death rate among all states.

{\centering \includegraphics{paper_files/figure-pdf/fig-deaths-rates-votes-1.pdf}
